### 2016-08-01 doc/publications/plos16/plos16.tex
many changes. added tableofcontents to help organize the paper

\end{flushleft}
% Please keep the abstract below 300 words

\section*{Notes}
\begin{itemize}
\item   I think the structure of the first section should be
  \begin{itemize}
  \item  convergence results from Biau 2012
  \item consider  \cite{Segal.2004} examples
  \item simulations reproduced from Genuer et al 2010 (no -- we need classification examples)
  \item check Piotr's simulation
  \item run sparkML or Cursed forest on the  simulation data 
  \item we extend the simulations to many more variables 
  \item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
    \ref{table:smaller_data} -- On it, Aidan
  \item variable  selection may succeed where accurate prediction does not
  \item discussion of the effect of the ntree and mtry parameters for variable selection with many noise variables
  \item   perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form (step, ramp) we can recover
    the significant variables in the presence of a very large number of noise variables
  \item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic
    functions. So we have some hope.
  \end{itemize}
\item check out 
  \begin{itemize}
  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). (done)
  \item Goldstein et al. (2011)
  \item Chen (2012)  (done)
  \item VSURF and  varSelRF, (doee) recursive feature (forwards backwards) using variable importance
  \item ranger package
  \item  r2VIM: A new variable selection method for random forests in
    genome-wide association studies
  \item CloudForest: A Scalable and Efficient Random Forest Implementation for Biological Data
  \item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this? (no, leave it for now)
  \end{itemize}
\item other things to do
  \begin{itemize}
  \item what is the problem with bagging? Do we need to redo the thousand genome stuff?
  \end{itemize}
\end{itemize}

\section*{Abstract}
The abstract.

@@

\section*{Introduction}

The digital revolution is seeing a dramatic increase in data collected about every aspect of 
life~\cite{Loebbecke2015}.  These dataset not only grow vertically by capturing more and more events but also
horizontally by capturing more information about these events.  The challenge of big and `wide' data is especially
pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
billion basepairs of the human genome.  While a large fraction of this data is identical between individuals, \marginpar{base-pairs}
identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].

For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  The richness of
this data can be utilised in other datasets as well by filling in or ``imputing" the information at genomic positions
previously unobserved due to the less detailed resolution of older array technology [citation about imputation].  This
means that the large number of samples for which genome-wide association (GWA) data is available~\cite{Welter2013} can
be extended horizontally by imputing the previously unobserved genotypes.  This provides the potential to generate
datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating
modern compute paradigms to deal with these challenges.

@@

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
supervised learning, i.e. given a phenotype of clinical label determine the most associated features.  One of the
simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
This is known as the `curse of dimensionality'~\cite{Bauer2014}.  
\marginpar{is this the reference we want for the ``curse of dimensionality''  Bellman (1961) is a more ususl reference}

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are scale invariant;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to useless predictors;
  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

\subsection*{RF}

@@

The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
the $i=1,..,Q$ classes
\begin{equation*}
{\bar {h}}=  \arg \max_i \left(\sum_b I(h(x;\theta_b)=i)\right).
\end{equation*}

Random forests are a variance reduction technique. They take a tree models, which have a low bias but a high variance, and
by combining them reduce the variance. The prediction error is the sum of the variance and the bias squared.

The average prediction error for an individual tree $h(X; \theta)$ is
\begin{equation}
PE_t = E_\theta E_{X,Y} (Y-h(X; \theta))^2.
\end{equation}
Assume that, for all  $\theta$, the tree is unbiased, i.e., $EY= E_X h(X; \theta)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;\theta))$ and $(Y-h(X;\theta^\prime))$ for independent $\theta,
\theta^\prime$.  
So what is required for  accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.

The benefits of random forests models are
  \begin{itemize}
  \item they give an out-of-bag estimate of model accurace;
  \item they provide a measure of variable importance;
  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
    variables. This would be difficult as the trees as split on random subsets of variables).
  \end{itemize}

Since the original implmentation of random forests there have been a number of developments leading to a better understanding of
the algorithm. 
\begin{itemize}
\item random forests are resistent against overfitting. However it is not true that they will not overfit; see \cite{Segal.2004}
  where an example of RF overfitting is  given, caused by the fact that successive trees were correlated;
\item \cite{Strobl.et.al.2007} demonstrates that the RF algorithm is subject to a bias in variable selection via two mechanisms:
  \begin{enumerate}
  \item variable scale (for $X_I$ continuous) and number of levels (for $X_i$ discrete) inroduces a bias. A uninformative  variable with a large
    number of levels may be selected over a more informative  variable with fewer levels;
  \item bagging introduces a bias, but sampleing about $(1- 1/e) \approx 0.632$  of the data without 
  \end{enumerate}
\end{itemize}

\cite{Strobl.et.al.2007} introduce changes to the RF algorithm to fix 1) and 2). 
However the level effect will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
However, there are many examples, (particularly in genomics) whre we have data sets that are
\begin{itemize}
\item very wide
\item all variables have the same number of levels (measures at different points along the genome)
\item the hypothesied functional relationship is of a form amenable to a DT.
\end{itemize}

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker.
\marginpar{say something about the rate of convergence}

@@

@@

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis"

\subsubsection*{variable selection}
\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using the
importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic data.
\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of RF to handle GWAS data. Waht do we nned to do in this area}

\section*{Methods}

\subsection*{VariantSpark.}
Traditionally, VariantSpark made use of Spark's machine learning algrorithms `Spark ML'. These algorithms require data in a `DataFrame',
similar to DataFrames in R.

@@

Cursed Forest uses:
\begin{itemize}
\item an ``information gained'' criteria for splitting.  $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
  \ldots, Q$ in a set. The  entropy of the set is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i$
The expected information gain is the change in information entropy $H$ from a prior state $T$ to a state that takes some
information $a$ as given: 
\[ IG(T,a) = H(T) - H(T|a) \]
In this case, the  the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
\item variable importance is the sum of IG over all notes and all trees for that variable
\item the model of parallelism is different to the ``foreach'' method -- we need to describe it
\end{itemize}

% Results and Discussion can be combined.
% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}. 
\section*{Results and Discussion}

@@

\marginpar{These are regression examples, we nned some classification examples}

@@

\subsection*{CursedForest processes genome-wide variant data}
%\label{comp}
%Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and
%time/mem)

@@

\subsection*{Parameter settings}
We consider the parameter setting for the RF algorithm. We use the R terms from the package RandomForests
\cite{Liaw.and.Weiner.2002} which incoporates the original Fortran code by Brieman and Cutler. We incprporate the 
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience.

\begin{itemize}
\item \texttt{ntree} the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
  \marginpar{can we do proxmiity?}  \cite{Liaw.and.Weiner.2002} suggest that a high \texttt{ntree} is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a RF model to have a very poor
  fit, and be useless for prediction, and still have a useful ranking of variable importance.
\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decison
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \texttt{mtry} may give
  better performance.
\item  For classification problems where the class frequencies are extremely unbalanced it may be necessary to change the prediction
  rule \marginpar{can we do this}
\item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark.ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
  example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.

@@

%\subsection*{James' dataset}
%Comparing CursedForest with feature selection + linear model (performance and time/mem)

\subsection*{Conclusion}

\section*{Supporting Information}

@@

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\tableofcontents
\clearpage
\section{Notes}
\begin{itemize}
\item   I think the structure of the first section should be
  \begin{itemize}
  \item  convergence results from Biau 2012
  \item consider  \cite{Segal.2004} examples
  \item simulations reproduced from Genuer et al 2010 (no -- we need classification examples)
  \item check Piotr's simulation
  \item run sparkML or Cursed forest on the  simulation data 
  \item we extend the simulations to many more variables 
  \item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
    \ref{table:smaller_data} -- On it, Aidan
  \item variable  selection may succeed where accurate prediction does not
  \item discussion of the effect of the ntree and mtry parameters for variable selection with many noise variables
  \item   perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form (step, ramp) we can recover
    the significant variables in the presence of a very large number of noise variables
  \item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic
    functions. So we have some hope.
  \end{itemize}
\item check out 
  \begin{itemize}
  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). (done)
  \item Goldstein et al. (2011)
  \item Chen (2012)  (done)
  \item VSURF and  varSelRF, (doee) recursive feature (forwards backwards) using variable importance
  \item ranger package
  \item  r2VIM: A new variable selection method for random forests in
    genome-wide association studies
  \item CloudForest: A Scalable and Efficient Random Forest Implementation for Biological Data
  \item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this? (no, leave it for now)
  \end{itemize}
\item other things to do
  \begin{itemize}
  \item what is the problem with bagging? Do we need to redo the thousand genome stuff?
  \end{itemize}
\end{itemize}

\clearpage

\section{Abstract}
The abstract.

@@

\section{Introduction}

The digital revolution is seeing a dramatic increase in data collected about every aspect of 
life~\cite{Loebbecke2015}.  These data sets are not only growing vertically by capturing more and more events but also
horizontally by capturing more information about these events.  The challenge of big and `wide' data is especially
pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
billion basepairs of the human genome.  While a large fraction of this data is identical between individuals, \marginpar{base-pairs}
identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].

For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  The richness of
this data can be utilised in other datasets as well by filling in or ``imputing" the information at genomic positions
previously unobserved due to the less detailed resolution of older array technology [citation about imputation].
\marginpar{citations missing} 
This means that the large number of samples for which genome-wide association (GWA) data
is available~\cite{Welter2013} can be extended horizontally by imputing the previously unobserved genotypes.  This
provides the potential to generate datasets of hundreds of thousands of individuals with millions of variants,
highlighting the need for incorporating modern compute paradigms to deal with these challenges.

@@

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
supervised learning, i.e. given a phenotype or clinical label determine the most associated features.  One of the
simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
This is known as the `curse of dimensionality'~\cite{Bauer2014}.  
\marginpar{is this the reference we want for the ``curse of dimensionality''  Bellman (1961) is a more usual reference}

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are invariant to monotonic scaling of the data;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to useless predictors;
  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

\subsection{RF}

@@

The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}=  \arg \max_q \left(\sum_b I(h(x;\theta_b)=q)\right).
\end{equation*}

Random forests are a variance reduction technique. They take unpruned tree models, which have a low bias but a high
variance, and by combining them reduce the variance. The prediction error is the sum of the variance and the bias
squared.

The average prediction error for an individual tree $h(X; \theta)$ is
\begin{equation}
PE_t = E_\theta E_{X,Y} (Y-h(X; \theta))^2.
\end{equation}
Assume that, for all  $\theta$, the tree is unbiased, i.e., $EY= E_X h(X; \theta)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;\theta))$ and $(Y-h(X;\theta^\prime))$ for independent $\theta,
\theta^\prime$.  
So what is required for  accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
on  bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.

The benefits of random forests models are
  \begin{itemize}
  \item they give an out-of-bag estimate of model accuracy, by using the samples not selected for the bootstrapped sample
    as an independent test set for each tree;
  \item they provide a measure of variable importance, for example by summing the number of times a variable is used to
    split a node, over all nodes and all trees. More complex measures of variable importance are also used;
  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
    variables. This would be difficult as the trees as split on random subsets of variables).
  \end{itemize}

Since the original implementation of random forests there have been a number of developments leading to a better understanding of
the algorithm. 
\begin{itemize}
\item random forests are resistant against overfitting. However it is not true that they will not overfit; see \cite{Segal.2004}
  where an example of RF overfitting is  given, caused by the fact that successive trees were correlated;
\item \cite{Strobl.et.al.2007} demonstrates that the RF algorithm is subject to a bias in variable selection via two mechanisms:
  \begin{enumerate}
  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a bias. A uninformative  variable with a large
    number of levels may be selected over a more informative  variable with fewer levels;
  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$  of the data without replacement seems to be an
    effective stragey for removing the problem.
  \end{enumerate}
\end{itemize}

\cite{Strobl.et.al.2007} introduce changes to the RF algorithm to fix 1) and 2). 
However the level effect will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
However, there are many examples, (particularly in genomics) where we have data sets that are
\begin{itemize}
\item very wide
\item all variables have the same number of levels (measures at different points along the genome)
\item the hypothesized functional relationship is of a form amenable to a decision tree.
\end{itemize}

\subsubsection{convergence}
Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker.
\marginpar{say something about the rate of convergence}

@@

\subsubsection{variable selection}
\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using the
importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic data.
\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of RF to handle GWAS data. What do we need to do in this area}

\section{Methods}

\subsection{VariantSpark.}
Traditionally, VariantSpark made use of Spark's machine learning algorithms `Spark ML'. These algorithms require data in a `DataFrame',
similar to DataFrames in R.

@@

@@

The current implementation of Cursed Forest uses:
\begin{itemize}
\item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
  \ldots, Q$ at a node. The  entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i$
The expected information gain is the change in entropy $H$ from a prior state $T$ to a state that takes some
information $a$ as given: 
\[ IG(T,a) = H(T) - H(T|a). \]
In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
\item variable importance is the sum of IG over all notes and all trees for that variable
\end{itemize}

\section{Results and Discussion}

@@

\marginpar{These are regression examples, we need some classification examples}

@@

\subsection{CursedForest processes genome-wide variant data}
%\label{comp}
%Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and
%time/mem)

@@

\subsection{Parameter settings}
We consider the parameter setting for the RF algorithm. We use the R terms from the package RandomForests
\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler. We incorporate the 
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience.

\begin{itemize}
\item \texttt{ntree} the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
  \marginpar{can we do proximity?}  \cite{Liaw.and.Weiner.2002} suggest that a high \texttt{ntree} is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a RF model to have a very poor
  fit, and be useless for prediction, and still have a useful ranking of variable importance.
\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \texttt{mtry} may give
  better performance.
\item  For classification problems where the class frequencies are extremely unbalanced it may be necessary to change the prediction
  rule \marginpar{can we do this}
\item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark.ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
  example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.

@@

%\subsection{James' dataset}
%Comparing CursedForest with feature selection + linear model (performance and time/mem)

\subsection{Conclusion}

\section{Supporting Information}

@@

% Results and Discussion can be combined.
% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
